---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
#quotes

x <- "a string"
x
y <- 'also a string'
y
z <- "this is a 'string' too"
z
w <- 'this is also a "string"'
w


# cat(), print()
print(x)
cat(x)

cat(x, file = "cat_example.txt") #prints it as a text file

sample.ttest <- t.test(rnorm(20, 0, 1), rnorm(20, 2, 1))
sample.ttest
print(sample.ttest)

printed.ttest <- print(sample.ttest) 

# special characters: \t, \n, "escaping"

x <- "here is a cool result: 10"
cat(x)
y <- "here's a cool result:\10"
y

cat("some\ntest")
cat("some\\ntext") #tells r to treat the 2nd slash as an actual slash
# cat("some\"text\\"")


# nchar()

nchar("hello") #number of characters
length("hello") #number of words

# paste(), strsplit()

paste("hello", "there")

paste("hello", "there", sep = "") #means that there is no space between the two pasted words
x <- c("here", "are", "some", "strings")
x

paste("prefix", x) #puts prefex infront of each "x" term
paste("prefix", x, sep = "") #prefix pasted to each word with no space

paste(x, collapse = " ") #collapse is another term for sep i think

# stringr package

x
y <- paste(x, "suffix", sep = "_")
y
w <- paste(x, "suffix", sep = ".")
w

strsplit(y, split = "_") #splits them into different lines.

strsplit(w, split = ".") #splitting on every character
strsplit(w, split = "\\.") #splits them into different lines.


# grep(), grepl(), gsub()
x
grep("here", x) #first put the string you are looking for, and then the location where its embeded
grep("e", x) #says that it found the letter e in the 1, 2,  and 3rd strings within "x" vector

grepl("here", x) #use this for logicals. get the true/false answers. can do a txt search 
                  #any time we find these characters report a true/false.

gsub("e", "E", x) #search and replace function. give a thing to search, then a thing to replace, then the vector
gsub("e", "", x)

longer.thing <- c("10f5675", "59j4957", "30k4934")

gsub("\\d+\\w(\\d+)", "\\1", longer.thing) #the number refers to the specific parenthesis

gsub("(\\d+)(\\w)(\\d+)", "\\2", longer.thing) #now its calling the 2nd parens (the letter value)




```

# Factors
  - Actually numbers under the hood VIP VIP VIP
  - Act like numbers + labels VIP VIP VIP
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()
#behave like they are numbers with labels. they have levels. duh.
#strings will often come in as factors. if you dont want that, you can define how the string is read in

x <- c(rep("a", 20), rep("b", 20), rep("c", 20))
x
summary(x)
x.factor <- factor(x)


# levels()
levels(x.factor) #we can get the levels. 

# summary()
summary(x.factor) #gives you the nubmer of cases

# coersion

# subsetting
sleep
mysleep <- sleep
summary(mysleep) #col 1 is min/max; group tells us 10 cases of 1 and 10 cases of 2
class(mysleep$group) #tell us if group is considered a factor
levels(mysleep$group) #tells us the number of levels

# droplevels
mysleep.group1 <- mysleep[mysleep$group == "1", ]
summary(mysleep.group1) #have all the data for group 1 only

mysleep.group1 <- droplevels(mysleep.group1) #directly changing this object to say
summary(mysleep.group1) #got rid of any level of a factor without data, no longer bugs you about it.

# changing levels (labels); change "onh" to "1"
mysleep <- sleep
levels(mysleep$group)

levels(mysleep$group) <- c("pre", "post") #this order matters. level 1 is space 1.

levels(mysleep$group)
summary(mysleep) #now it defines it as pre/post in the data output.

head(mysleep)

mysleep.bad <- sleep
levels(mysleep.bad$group) <- c("2", "1")
summary(mysleep.bad$group)


colnames(mysleep) <- c("extra sleep", "treatment group", "subject")
head(mysleep)

# re-ordering
mysleep <- sleep
levels(mysleep$group) <- c("pre", "post")
summary(mysleep)
#now was want to move post to level 2 and pre to level 1
# levels(mysleep$group) <- c("post", "pre") #changes data labels, not order of levels. screws up data
# levels(mysleep$group)

#might do this to change the referent level
mysleep$group2 <- mysleep$group #assign value to column that doesnt exhist, it will create the column
head(mysleep)
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre"))
summary(mysleep)
head(mysleep)
tail(mysleep)


```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value
mysleep$group3 <- mysleep$group #can just make a new column, dont need to do something special
mysleep$experiment <- 1 #gave the whole column a single value "1"
summary(mysleep)

#ifelse()
mysleep$sleeplevel <- ifelse(mysleep$extra >= 1, "high", "low") #true = high, false = low
summary(mysleep)
mysleep$sleeplevel <- as.factor(ifelse(mysleep$extra >= 1, "high", "low"))
summary(mysleep)
mysleep$sleeplevel <- factor(ifelse(mysleep$extra >= 1, "high", "low"), levels = c("low", "high"))
summary(mysleep)

mysleep$sleeplevel <- ifelse(mysleep$extra >=1, "high", ifelse(mysleep$extra < 1, "low", NA))
summary(mysleep)

# using subsetting

```

## Using xtabs() to check coding/cross-tabs

```{r}
xtabs(~ group + group2, data = mysleep) #allows us to double check that if its pre in the data it isnt also post
xtabs(~ sleeplevel + extra, data = mysleep)
xtabs(~ extra + sleeplevel, data = mysleep) #this is a frequency plot; after 2 levels it gets to a mess.

```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}
#missing data - could be a value but i dont know what it is; there could have been a value but i dont know it
#NAs means there isnt an existing value
#cor() can tell it to deal with NAs by only running the data without the NAs.
#ignore the NAs, need to tell it for each function "na.rm = true"
```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}
sqrt(-1)
```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

#is.X functions tell you if something is true/false whatehver you want.
```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# PRACTICE: #7

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# PRACTICE: #8 -- #11

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details



# PRACTICE: #12 -- #14


